---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, # TODO: delete this line when package has functional examples
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# checkthat

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/checkthat)](https://CRAN.R-project.org/package=checkthat)
[![R-CMD-check](https://github.com/iancero/checkthat/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/iancero/checkthat/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/iancero/checkthat/branch/main/graph/badge.svg)](https://app.codecov.io/gh/iancero/checkthat?branch=main)
<!-- badges: end -->

The goal of checkthat is to integrate testing into your data manipulation pipelines.

## Installation

You can install the development version of checkthat like so:

```{r, eval=FALSE}
devtools::install_github('iancero/checkthat')
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r}
stop('Banana')
```


```{r example}
library(checkthat)

df <- mtcars
df$id <- rownames(mtcars)
rownames(df) <- NULL

df <- df[, c('id', names(df)[names(df) != 'id'])]

mtcars |> 
  check_that(
    all(cyl >= 4),
    any(cyl < 8),
    
    some(cyl < 8, at_least = 5),
    some(disp > 200, at_least = .10, less_than = 75),
    
    whenever(
      is_observed = cyl == 6, 
      then_expect = mpg < 50.0),
    
    specifically(case = 2, mpg == 21, cyl < 8),
    specifically(case = id == 'Mazda RX4 Wag', hp == 110)
  )

df2

```




## Alternatives

There are two pre-existing packages that served as inspiration for checkthat. They are both quite good and, depending on your use case, might be a better choice for you.

- [**testthat**](https://testthat.r-lib.org/) implements **unit testing for packages** and is currently the most popular testing package for R. If your goal is to develop a package - rather than conduct a data analysis - then testthat will be a much better choice than checkthat.

- [**testdat**](https://socialresearchcentre.github.io/testdat/index.html) package is inspired by testhat and - like checkthat - also implements **unit testing for data manipulation**. It is different from checkthat in two ways. First, it is a more mature package (i.e., less buggy, more feature rich) than checkthat. Second, whereas checkthat is designed to integrate tests into data manipulation pipelines, testdat is designed to place tests in separate commands or files. Thus, as checkthat matures, the difference between the two packages will become largely stylistic.






















