---
title: "checkthat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{checkthat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(checkthat)
```

The **checkthat** philosophy is that you already perform good data checks and you should keep doing it. But those checks would be even better if they lived in the code, rather than in your head. Checkthat therefore provides functions that closely resemble the checks you already do by hand or by eye, so that it is easy for you to also express them in code as you go.

## Basic usage 

Checkthat's main function is `check_that(.data, ...)`, which takes a dataframe as its first argument, followed by any number of assertions you want to check for that dataframe.

When all checks pass, you get a brief message confirming that's the case.

```{r}
library(checkthat)

mtcars |>
  check_that(
    all(cyl > 2),
    !any(is.na(mpg))
  )
```

When at least one check fails, `check_that()` throws an error, halting the potentially risky execution of subsequent code. It then gives you get a detailed breakdown of what the outcome was for each test.

```{r, echo=FALSE}
set.seed(123456)
```

```{r, error=TRUE}
mtcars |>
  check_that(
    all(cyl > 2),
    any(mpg > 35)
  )
```

The `check_that()` function is designed to work with both base R's existing logical functions (e.g., `all()`, `any()`), as well it's own set of more special helper functions. Theses helper functions are designed to be both readable and to mirror in code what you already do manually by eye-balling a dataset.

```{r}
mtcars |>
  check_that(
    some_of(cyl > 4, at_least = .30, at_most = 25),
    whenever(wt < 3, then_expect = mpg > 19),
    for_case(2, mpg == 21, hp == 110)
  )
```


## Tidyverse pipelines

The `check_that()` function always invisibly returns the same `.data` you gave it (always unmodified). This allows you to easily integrate it directly into your data manipulation pipelines.

```{r}
new_mtcars <- mtcars |>
  select(mpg) |>
  mutate(km_per_litre = 0.425 * mpg) |>
  check_that(max(km_per_litre) < 15)

head(new_mtcars)
```

## Checking a pipeline at multiple points

Because it returns the same dataframe it received, `check_that()` can also be used at multiple points in a single pipeline. That way, you can check that multi-step processes are unfolding according to plan. This is be especially important for data tasks that are sensitive to the order of operations, or for checks on intermediate data that wont be available at the end.

Consider a surprisingly tricky example. Imagine we wanted to (1) create a factor variable (`type`) designating cars as either small (`"sm"`) or large (`"lg"`) based on their weight (`wt`). Further imagine that we then (2) planned to filter in only the small cars and (3) calculate their mean `mpg` as our `desired_mpg`. This value might then be used to inform a personal purchase decision or perhaps to establish an industry benchmark for a manufacturer.

The resulting data pipeline should be simple, but let's use `check_that()` at multiple points to be safe. 

1. We wont have access to the `wt` variable at the end of the pipeline. So, right after we use `wt` to compute `type`, we immediately check that all the weights in the `"sm"` group are less than those in the `"lg"` group, as intended.
2. At the end, we check that our `desired_mpg` is within a plausible range.

Here, the first check throws an error and stopS the pipeline. It also saves us from an inaccurate `desired_mpg` that the second check would not have caught.

```{r, error=TRUE}
mtcars |>
  mutate(type = factor(wt < 3, labels = c("sm", "lg"), ordered = TRUE)) |>
  check_that(max(wt[type == "sm"]) <= min(wt[type == "lg"])) |>
  filter(type == "sm") |>
  summarise(desired_mpg = mean(mpg)) |>
  check_that(desired_mpg > 15)
```

What happened? A quick reading of `factor(wt < 3, labels = c("sm", "lg"), ordered = TRUE)` seems like it would correctly assign cars to the correct group. However, the labels are out of order in the function call.^[The `factor()` function maps `wt < 3 == FALSE --> 1` and `TRUE --> 2` in this case because `FALSE < TRUE`. So, when `wt < 3 == TRUE`, `factor(..., labels = c("sm", "lg"))` will assign the 2nd label (mistakenly, from our perspective).] As a result, the heavy cars are mistakenly labelled `"sm"` and vice-versa.^[Full disclosure, I made this mistake when preparing this example, and it was the `check_that()` function that pointed it out to me. So, I decided to include it.]

Importantly, this mistake (a) would have given us an erroneously low `desired_mpg` and (b) would have gone undetected by our final `check_that(desired_mpg > 15)`. It was a call to `check_that()` earlier in the pipeline that caught the error and prevented us from drawing an bad conclusion about our data later on.

## Helper functions

Checkthat's philosophy is your existing data checks by eye are probably already good. Their only major problem is that they live in your head and not in your code. So, checkthat provides a range of helper functions to work alongside base R's existing collection (e.g., `all()`, `any()`). These include both some basic and more special varieties.

### Basic helpers

The most basic helpers are just syntactic sugar around R's existing comparison operators: `=`, `<`, `<=`, `>`, `>=`. Each of them takes a logical vector as its first argument and requires you to specify a proportion (`p`) or count (`n`) of those values that must be true.

```{r}
mtcars |>
  check_that(
    at_least(mpg < 35, p = .95),
    more_than(hp == 110, n = 2),
    exactly_equal(cyl == 6, n = 7),
    less_than(wt > 3, p = .75),
    at_most(is.na(mpg), n = 3),
  )
```

### Special helpers

The remaining helpers include `some_of()`, `whenever()`, and `for_case()` and are more flexible than their basic counterparts. They're optimized for the kind of semi-approximate data checking you are likely already doing by eye. 

For most people, this involves a general sense of what most of the data should look like most of the time, but not exact knowledge of specific proportions or counts. For example, you might have good reason to think `some_of()` the `cyl` values should be greater than 4, but you don't know exactly how many. However, you do know it should probably be `at_least` 30%, but `at_most` 25 total cases in your dataset. Anything outside that range would be implausible and so you want to guard it with `check_that()`. 

```{r}
mtcars |>
  check_that(
    some_of(cyl > 4, at_least = .30, at_most = 25),
    whenever(is_observed = wt < 3, then_expect = mpg > 19),
    for_case(2, mpg == 21, hp == 110)
  )
```

Just like unit tests for production code, the tests created with these special helper functions will be technically imperfect and leave some (possibly important) scenarios addressed. After all, there's a big range of possibilities between `at_least = .30` and `at_most = 25`, and some of them might involve an undetected data problem.

However, checkthat takes the position that imperfect tests are still valuable informative and you should be able to take advantage of them. For example, if you have reasons to be concerned about the data in your column crossing the `at_most = 25`, you should be able to quickly and easily write that test with a combination of `check_that()` and `some_of()`.

Moreover, a world of *no tests at all* is much worse than a world of *some tests that fail to cover every case*. With that in mind, checkthat's special helper functions are designed to bring you from *not writing down any tests in your code* to *quickly and easily coding the tests you already do by eye*.

## Checking the whole dataframe

In addition to concerns about the individual rows or columns in your data, you may also want to perform checks on the entire dataframe in question. For those cases, `check_that()` provides the `.d` pronoun, which works similarly to `.x` in the **purrr** package.

In short, `.d` is a copy of the data you provided to `check_that()`, which you can use to write checks about the whole dataset.

```{r}
mtcars |>
  check_that(
    nrow(.d) > 10,
    "mpg" %in% names(.d)
  )
```

This is especially useful for operations that could change the shape of your dataset (e.g., pivots, nests, joins). In the case of pivoting, you might want to check that the dataset have the correct anticipated dimensions.

```{r}
mtcars |>
  check_that(ncol(.d) == 11, nrow(.d) == 32) |> # original dimensions
  pivot_longer(
    cols = everything(),
    names_to = "name",
    values_to = "values"
  ) |>
  check_that(ncol(.d) == 2, nrow(.d) == 32 * 11) # check that cols became rows
```

After a join, you may want to check that there is a new column in the expected location, but also that there are no unanticipated new rows.

```{r}
cyl_ratings_df <- data.frame(cyl = c(4, 6, 8), group = c("A", "B", "C"))

mtcars |>
  left_join(cyl_ratings_df, by = "cyl") |>
  check_that(
    ncol(.d) == 12, # check that there's one new column
    names(.d)[length(names(.d))] == "group", # check new column is "group"
    nrow(.d) == 32 # check that no new rows
  )
```
